### 2017-07-05 paper/synthesis.tex
re-shuffled the proof of the main theorem a bit

In this section we describe a polynomial-time algorithm for synthesizing linear $\eps$-LexRSM maps in affine probabilistic programs supported by a given linear invariant map $\inv$. The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{xxx}. Hence, we provide only a high-level description, focusing on new aspects of our adaptation.

The main idea is to iteratively synthesize 1-dimensional linear expression map that $1$-rank a subset of generalized transitions. These maps form the individual components of the sought-after $1$-LinLexRSM map. In each iteration, we start with a set $U$ of the yet-unranked generalized transitions. We seek a 1-dimensional LEM which ranks the maximal number of elements if $U$, and is unaffected by the remaining elements of $U$ (here by ranking a generalized transition $\tilde\tau$ we mean ranking it from each configuration $(\loc,\vec{x})$, where $\loc$ is the source location of $\tilde\tau$ and $\vec{x}\in\inv(\loc)$). If no 1-dimensional LEM that would rank at least one element in $U$ exists, then there is no LinLexRSM map for the program. Otherwise, we remove the newly ranked elements from $U$ and continue into the next iteration, until $U$ becomes empty. The process is summarized in Algorithm~\ref{xxx}.

Hence, the main computational task of the algorithm is to check, for a given set of generalized transition $U$, whether there exists a 1-dimensional LEM $\lem$ such that:
\begin{compactenum}
\item for each location $\loc\in\locs$ and all $\vec{x}\in \inv(\loc)$ it holds $\lem(\loc,\vec{x})\geq 0$;
\item for each $\tilde{\tau}\in U$ and  each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$ we have that $\tilde{\tau}$ is unaffected by $\lem$ from $(\loc,\vec{x})$; and
\item there is $\tilde{\tau}\in U$ that is $1$-ranked by $\lem$, from each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$; we then say that $\lem$ ranks $\tilde{\tau}$ w.r.t. $\inv$.
\end{compactenum}
  Moreover, if such an LEM $\lem$ exists, the algorithm has to find one that maximizes the number of gen. transitions in $U$ ranked by it. Both these tasks can be accomplished by the standard  method of linear constraints based on the use of Farkas's lemma, which was widely use for synthesis of termination proofs in both probabilistic and non-probabilistic programs~\cite{xxx}. That is, the algorithm first constructs, for each location $\loc$ a \emph{template} for $\lem$, i.e. an expression of the form $a_1^{\loc}x_1 + \cdots + a_{|\pvars|}^{\loc}x_{|\pvars|} = b^{\loc} $, where $x_1,\dots,x_{|\pvars|}$ are program variables and $a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc} $ are yet unknown coefficients. That is, supplying concrete values for all the unknown coefficients yields an LEM. Now the conditions (1) and (2) above can be expressed using linear constraints on the coefficients. More precisely, using the construction provided e.g. in~\cite{xxx} (which includes a use of the Farkas's lemma) we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints $\linsystem_{\tilde{\tau}}$ over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}\cup F$, where $F$ is the set of fresh variables (not appearing in any template) and $\eps_{\tilde\tau} $ is constrained to be non-negative. Each solution of the system $\linsystem_{\tilde{\tau}}$ yields a LEM which satisfies the constraints (1) and (2) for $\tilde{\tau}$. Moreover, each solution of $\linsystem_{\tilde{\tau}}$ yields a LEM which $\eps_{\tilde{\tau}}$-ranks $\tilde{\tau}$. To find a LEM which satisfies all constraints (1)--(3) as well as maximizes the number of 1-ranked elements of $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
  \begin{align*}
\text{maximize }  &\sum_{\tilde{\tau} \in U} \eps_{\tilde\tau} \text{ subject to constraints}\\[0.3cm]
&\linsystem_{\tilde{\tau}}\,;\quad\quad \quad\quad\tilde{\tau}\in U\\
&0 \leq \eps_{\tilde{\tau}} \leq 1\,; \quad \tilde{\tau}\in U
  \end{align*}

@@

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} follow by arguments identical to those presented in~\cite{xxx}. 
\begin{theorem}
\label{thm:algo}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. The algorithm runs in time polynomial in size of 
	$\program$ and $\inv$. \textbf{[PETR: POLYHEDRAL GUARDS]}
\end{theorem}
\begin{proof}[Proof (sketch)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$. Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.

Proving completeness is more intricate; as pointed out in~\cite{xxx}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs. In~\cite{xxx} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine. The same geometric properties hold for our \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), the result is easily transferable.
\end{proof}

\

>>>>

In this section we describe a polynomial-time algorithm for synthesizing linear $\eps$-LexRSM maps in affine probabilistic programs supported by a given linear invariant map $\inv$. The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{ADFG10:lexicographic}. Hence, we provide only a high-level description, focusing on new aspects of our adaptation.

The main idea is to iteratively synthesize 1-dimensional linear expression map that $1$-rank a subset of generalized transitions. These maps form the individual components of the sought-after $1$-LinLexRSM map. In each iteration, we start with a set $U$ of the yet-unranked generalized transitions. We seek a 1-dimensional LEM which ranks the maximal number of elements if $U$, and is unaffected by the remaining elements of $U$ (here by ranking a generalized transition $\tilde\tau$ we mean ranking it from each configuration $(\loc,\vec{x})$, where $\loc$ is the source location of $\tilde\tau$ and $\vec{x}\in\inv(\loc)$). If no 1-dimensional LEM that would rank at least one element in $U$ exists, then there is no LinLexRSM map for the program. Otherwise, we remove the newly ranked elements from $U$ and continue into the next iteration, until $U$ becomes empty. The process is summarized in Algorithm~\ref{algo:linlexrsm}.

Hence, the main computational task of the algorithm is to check, for a given set of generalized transition $U$, whether there exists a 1-dimensional LEM $\lem$ such that:
\begin{compactenum}
\item for each location $\loc\in\locs$ and all $\vec{x}\in \inv(\loc)$ it holds $\lem(\loc,\vec{x})\geq 0$;
\item for each $\tilde{\tau}\in U$ and  each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$ we have that $\tilde{\tau}$ is unaffected by $\lem$ from $(\loc,\vec{x})$; and
\item there is $\tilde{\tau}\in U$ that is $1$-ranked by $\lem$, from each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$; we then say that $\lem$ ranks $\tilde{\tau}$ w.r.t. $\inv$.
\end{compactenum}
  Moreover, if such an LEM $\lem$ exists, the algorithm has to find one that maximizes the number of gen. transitions in $U$ ranked by it. Both these tasks can be accomplished by the standard  method of linear constraints based on the use of Farkas's lemma, which was widely use for synthesis of termination proofs in both probabilistic and non-probabilistic programs~\cite{DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,SriramCAV,CFNH16:prob-termination}. That is, the algorithm first constructs, for each location $\loc$ a \emph{template} for $\lem$, i.e. an expression of the form $a_1^{\loc}x_1 + \cdots + a_{|\pvars|}^{\loc}x_{|\pvars|} = b^{\loc} $, where $x_1,\dots,x_{|\pvars|}$ are program variables and $a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc} $ are yet unknown coefficients. That is, supplying concrete values for all the unknown coefficients yields an LEM. Now the conditions (1) and (2) above can be expressed using linear constraints on the coefficients. More precisely, using the construction provided e.g. in~\cite{xxx} (which includes a use of the Farkas's lemma) we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints $\linsystem_{\tilde{\tau}}$ over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}\cup F$, where $F$ is the set of fresh variables (not appearing in any template) and $\eps_{\tilde\tau} $ is constrained to be non-negative. Each solution of the system $\linsystem_{\tilde{\tau}}$ yields a LEM which satisfies the constraints (1) and (2) for $\tilde{\tau}$. Moreover, each solution of $\linsystem_{\tilde{\tau}}$ yields a LEM which $\eps_{\tilde{\tau}}$-ranks $\tilde{\tau}$. To find a LEM which satisfies all constraints (1)--(3) as well as maximizes the number of 1-ranked elements of $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
  \begin{align*}
\text{maximize }  &\sum_{\tilde{\tau} \in U} \eps_{\tilde\tau} \text{ subject to constraints}\\[0.3cm]
&\linsystem_{\tilde{\tau}}\,;\quad\quad \quad\quad\tilde{\tau}\in U\\
&0 \leq \eps_{\tilde{\tau}} \leq 1\,; \quad \tilde{\tau}\in U
  \end{align*}

@@

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} follow by arguments identical to those presented in~\cite{xxx}. 
\begin{theorem}
\label{thm:algo}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. The algorithm runs in time polynomial in size of 
	$\program$ and $\inv$. \textbf{[PETR: POLYHEDRAL GUARDS]}
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$. Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.

Proving completeness is more intricate; as pointed out in~\cite{ADFG10:lexicographic}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs. In~\cite{ADFG10:lexicographic} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine. The same geometric properties hold for our generalization to \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), so the result is easily transferable.
\end{proof}

%\
